### 2016-06-24 publications/Journal/1-introduction.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
The selected synchronization protocol is transparent to the simulated model: users should merely determine, which protocol they wish to use.
Users who simulate a wide variety of models, with different ideal synchronization protocols, can simply run the same model with different synchronization protocols.

Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance.
Unlike PythonPDEVS dxex only supports multicore parallelism.

We implemented a model that, depending on a single parameter, changes its ideal synchronization protocol.
Dxex, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol. With dxex users can always opt to use the fastest protocol available.
To verify that our flexibility does not counter performance, we compare to adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS,DEVSSurvey}.

>>>>

This paper introduces DEVS-Ex-Machina\footnote{\url{https://bitbucket.org/bcardoen/devs-ex-machina}} (``dxex''), our simulation tool which offers multiple synchronization protocols: no synchronization (sequential execution), conservative synchronization, or optimistic synchronization.
The selected synchronization protocol is transparent to the simulated model: users should merely determine, which protocol they wish to use.
Users who simulate a wide variety of models, with different ideal synchronization protocols, can simply run the same model with different synchronization protocols.
We investigate in this paper how model allocation and uncertainty determine the choice between synchronization protocols. The synchronization overhead is demonstrated by reducing the computational load of a model to near zero. Dxex also offers a visualization of the simulation with in depth statistics on the synchronization algorithms. With all this information the modeller can make the best informed decision on which synchronization algorithm to use.

Our tool is based on PythonPDEVS, but implemented in C++11 for increased performance, using features from the new C++14 standard when possible.
Unlike PythonPDEVS dxex only supports multicore parallelism.

We implemented a model that, depending on a single parameter, changes its ideal synchronization protocol. We demonstrate using several models the factors influencing the performance under a given synhronization protocol.
Dxex, then, is used to compare simulation using exactly the same tool, but with a varying synchronization protocol. With dxex users can always opt to use the fastest protocol available.
To verify that our flexibility does not counter performance, we compare to adevs, currently one of the fastest \textsf{DEVS} simulation tools available~\cite{PythonPDEVS,DEVSSurvey}.

% TODO uncomment if completed
%Dxex offers online switching functionality that can be used by machine learning %techniques to automatically switch between the most optimal synchronzation protocol.

### 2016-06-24 publications/Journal/3-features.tex
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

\newcommand{\plotfraction}{0.85}
Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
Dxex is a C++14-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, while making some of its own additions.
So while the feature set is not too comparable, the architectural design, core simulation algorithm, and optimizations, are highly similar.

We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
In our case, a user can pass the simulation kernel the synchronization protocol to use for this model.
Our implementation in C++14 also allows for optimizations which were plainly impossible in an interpreted language.

Since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
This is due to dxex models being grafted on C++14, whereas PythonPDEVS models are grafted on Python.

@@

Reading and writing to shared memory is done through the use of the new C++14 synchronization primitives.
Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
C++14 further allows us to make the implementation portable, as well as more efficient: the compiler might know of optimizations specific to atomic variables.

@@

The main problem we encountered in our implementation is the aggressive use of memory.
Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
This made us switch to the use of thread-local (using \texttt{tcmalloc}) memory pools.
Again, we made use of specific new features of C++14, that were not available in Python, or even previous versions of the C++ language standard.

\subsection{Transparency}
We define simulation kernel transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
User should thus only provide one model, implemented in C++14, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
Switching between simulation kernels is as simple as altering the simulation termination time.
The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.

@@

In this section we highlight some of the extra features dxex offers to modellers.

@@

Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : json, XML, cell, console, ... . The user can implement his own outputpolicy and use it in our modular tracing design. The tracing functionality itself is implemented asynchronous and thread-safe. By overriding the output functions in the Message or State classes the user can generate model specific output for use with an existing tracing output policy. Tracing can be completely disabled to maximize simulation performance should this be desired (e.g. benchmarks). 

@@

As we will demonstrate in \ref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel. Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation. This allows us by using a single compilation parameter to change the allocation strategy per kernel type. This interface allows instrumentation of memory usage. Not only useful to detect memory leaks should this be a concern, it can also be of use to a model author to trace event creation in large models. A model author can rely on the default implementation as the fastest common configuration verified by the benchmarks.

>>>>

\newcommand{\plotfraction}{0.85}
Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
Dxex is a C++11-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, while making some of its own additions.
So while the feature set is not too comparable, the architectural design, core simulation algorithm, and optimizations, are highly similar.

We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
In our case, a user can pass the simulation kernel the synchronization protocol to use for this model.
Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language. Dxex will use new optimizations from the C++14 standard when possible.

Since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
This is due to dxex models being grafted on C++11, whereas PythonPDEVS models are grafted on Python.

@@

Reading and writing to shared memory is done through the use of the new C++11 synchronization primitives.
Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
C++11 further allows us to make the implementation portable, as well as more efficient: the compiler might know of optimizations specific to atomic variables or constant expressions which are heavily used in dxex.

@@

The main problem we encountered in our implementation is the aggressive use of memory.
Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
This made us switch to the use of thread-local (using \texttt{tcmalloc}) memory pools.
Again, we made use of specific new features of C++11, that were not available in Python, or even previous versions of the C++ language standard.

\subsection{Transparency}
We define simulation kernel transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
User should thus only provide one model, implemented in C++11, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
Switching between simulation kernels is as simple as altering the simulation termination time.
The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.

@@

We will enumerate some of the extra features dxex offers to modellers.

@@

Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : Json, XML, Cell, console, ... . The user can implement his own outputpolicy and use it in our modular tracing design. The tracing functionality itself is implemented asynchronous and thread-safe. By overriding the output functions in the Message or State classes the user can generate model specific output for use with an existing tracing outputpolicy. Tracing can be completely disabled to maximize simulation performance should this be desired (e.g. benchmarks). 

@@

As we will demonstrate in Section~\ref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel. Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation. This allows us by using a single compilation parameter to change the allocation strategy per kernel type. This interface allows instrumentation of memory usage. Not only useful to detect memory leaks should this be a concern, it can also be of use to a model author to trace event creation in large models. A model author can rely on the default implementation as the fastest common configuration verified by the benchmarks.

### 2016-06-24 publications/Journal/conclusion.tex
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

\subsection{Conclusion}
We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
We will summarize the key aspects in achieving good parallel performance here. % reword aspects syn
\subsubsection{Allocation}
Allocation of atomic models over kernels is critical to obtain a speedup. In the case of the Interconnect model it is not possible to allocate models without introducing a cyclic dependency between kernels with severe performance degradation as a result. 
Even when no cycles exist in the topology between kernels a good allocation scheme will minimize the number of connections between kernels and aim for a star topology as in PHoldTree with depth first allocation or a chain topology as in the Queue model. 
Dxex can optionally offer the user a visualization of the behavior of the kernels under a user supplied allocation scheme to allow for more insight. 
The same instrumentation could in future be used to optimize an existing allocation scheme.
\subsubsection{Uncertainty}
A simulation where future behaviour cannot be predicted is typically not suited for conservative simulation, but we have demonstrated that dxex's conservative kernel can still offer a non trivial speedup in this context. 
Optimistic is not constrained by this uncertainty and can offer a good speedup regardless of lookahead. 
PHOLD and PholdTree demonstrate that a single model can benefit from different synchronization protocols depending on a single parameter.
\subsubsection{Limits}
Dxex's conservative kernel is constrained by the amount of models it controls, as shown in the PHoldTree benchmark. This is especially the case when the minimal lookahead is $\epsilon$ requiring near constant polling of models for lookahead values. 
Optimistic is not constrained by the number of models it manages, but can quickly consume all available memory in a simulation. While this effect can be reduced with a thread aware allocation scheme, the underlying problem remains.

>>>>

\subsection{Conclusion}
We have shown that our contribution is invaluable for high performance simulation: depending on the expected behaviour, modellers can choose the most appropriate synchronization protocol.
We will summarize the key aspects in achieving good parallel performance here. % reword aspects syn
\subsubsection{Allocation}
Allocation of atomic models over kernels is critical to obtain a speedup. 
In the case of the Interconnect model it is not possible to allocate models without introducing a cyclic dependency between kernels with severe performance degradation as a result. 
Even when no cycles exist in the topology between kernels a good allocation scheme will minimize the number of connections between kernels and aim for a star topology as in PHoldTree with depth first allocation or a chain topology as in the Queue model. 
Dxex can optionally offer the user a visualization of the behaviour of the kernels under a user supplied allocation scheme to allow for more insight. 
The same instrumentation could in future be used to optimize an existing allocation scheme.
\subsubsection{Uncertainty}
A simulation where future behaviour cannot be predicted is typically not suited for conservative simulation, but we have demonstrated that dxex's conservative kernel can still offer a non trivial speedup in this context. 
Optimistic is not constrained by this uncertainty and can offer a good speedup regardless of lookahead. 
Phold and PholdTree demonstrate that a single model can benefit from different synchronization protocols depending on a single parameter.
\subsubsection{Limits}
Dxex's conservative kernel is constrained by the amount of models it controls, as shown in the PHoldTree benchmark. This is especially the case when the minimal lookahead is $\epsilon$ requiring near constant polling of models for lookahead values. 
Optimistic is not constrained by the number of models it manages, but can quickly consume all available memory in a simulation. While this effect can be reduced with a thread aware allocation scheme, the underlying problem remains.

### 2016-06-24 publications/Journal/parallel-performance.tex
First set of revisions and fixes.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

	\includegraphics[width=\modelfraction\columnwidth, height=8cm, keepaspectratio]{fig/queue_allocation.eps}
	\caption{Queue model (d=2, w=7, t=5000, random timeadvance) allocation and simulation trace across 4 kernels.}
	\label{fig:Queue_allocation}

@@

% The real life
\subsubsection{PholdTree}
%move to pholdtree
We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization. We will demonstrate that allocation is critical to achieve a parallel speedup and that the parameters of the model will give either conservative or optimistic an advantage.
\paragraph*{Allocation}\label{PholdTreeallocation}
The PholdTree benchmark can be configured to use 2 allocation schemes : breadth first and depth first. The breadth first allocation scheme will result in kernels that will form a dependency chain between them with multiple branches, much like in the Queue model. Such a linear dependency chain can result in a parallel speedup as we demonstrate with the Queue model, but this need not be true for all models as we will demonstrate in this section. A single kernel that has an unbalanced number of atomic models, models with unequal computation load in transition functions or activity will slow down the remainder of the chain. This effect is also apparent if the OS unfairly schedules the thread a kernel runs on. In conservative this will lead to excessive polling of the other kernels' eot values, whereas in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel. In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.

@@

Adevs' conservative implementation cannot handle the uncertainty (lack of non $\epsilon$ lookahead) here and is omitted from the comparison. Its parallel performance is two orders of magnitude slower than the sequential implementation. Upon investigation we conclude that this slowdown is caused by not implementing the \{begin/end\}lookahead() functions, which when not overridden in a model trigger an exception on each invocation. \\
This exception handling completely stalls the kernel. We do not implement this function in our PholdTree variant for adevs since this would force us to implement state saving in the model code and not use the provided state saving functionality in the kernel, as is done in dxex or PythonPDEVS's optimistic.\\
In dxex state saving is never required for a conservative simulation, this is handled transparently for the user who need not implement this behavior. 
By using the state saving technique inside the model code we feel we would no longer compare identical models across different synchronization algorithms, although we do not doubt adevs' increased performance when these functions are overridden.
\\

@@

% Conservative
Conservative has a more nuanced speedup behavior in this benchmark. We have detailed how a conservative kernel in dxex scales linear in the number of atomic models (if lookahead is $\epsilon$), this effect is more clearly visible here. \\
It is clear that as the fanout (n) of the model increases the kernel performance degrades rapidly. The intersection of the performance graph of each configuration is shifted to the right, the actual 1-speedup point increases as the kernelcount increases. For a d=4, n=5 configuration with 4 kernels each kernel has a load of ~1300 atomic models when it crosses the 1-speedup line. \\
An 8 kernel configuration with d=4 n=6 can sustain a load of ~2700 atomic models before it reaches that point. From the results we see that the 32-kernel configuration is not yet slowing down with the current parameters. \\
The modelcount is only a part of the explanation of this effect, as the kernelcount increases the amount of inter kernel messages will be split into more distinct sets which can be handled with more concurrency in dxex's architecture. 
That this effect is not always true can be observed from the d=4 n=3 32-kernel datapoint. Note that in this configuration 341 atomic models are distributed over 32 kernels, a relative low model load can more easily highlight synchronization overhead, even if allocation is optimal.\\
As with optimistic we do not include breadth first allocated benchmarks in this speedup plot since none of those achieve a speedup higher than 1.
\\
If we combine both to determine which synchronization protocol is more optimal we see in Figure \ref{fig:PholdTree_plot_weakall} that for these configurations conservative with 8 kernels is an ideal configuration up until n=5, where the modelcount starts to degrade performance. Optimistic with the same kernelcount is almost insensitive to the increase in modelcount and is therefore a more robust choice. Note how optimistic and conservative with 32 kernels at this point still have a reasonable speedup which is surprising given the synchronization overhead in such a large set of kernels.
\begin{figure}
	\center

@@

\subsection{Measuring Synchronization Overhead}
The simulation kernels have a non trivial overhead compared to a sequential kernel. In this section we will try to isolate this overhead and highlight the consequences this has for applying a synchronization protocol in real life models.
Profiling indicates that there are in the most trivial of models only 2 causes for computational load outside of the actual simulation : memory management and random number generation. 
\paragraph*{Memory management}\label{sec:4-subsec:overhead-pgraph:memory}
\begin{figure}
	\center

@@

A sequential simulation kernel can leverage the invariant that no allocated event will persist beyond a single time advance and use a arena-style allocator. Conservative and optimistic simulation need to use generic pool allocators. \\

@@

\end{figure}
% Analyse

>>>>

	\includegraphics[width=\modelfraction\columnwidth, height=8cm, keepaspectratio, angle=-90 ]{fig/queue_allocation.eps}
	\caption{Queue model (d=2, w=7, t=5000, random timeadvance) allocation and simulation trace across 4 kernels.}
	\label{fig:Queue_allocation}

@@

% The real life
\subsubsection{PholdTree}
%move to pholdtree
We further verify that our contribution fulfills our projected use case: a single model that can be tweaked to favor either conservative or optimistic synchronization. We will demonstrate that allocation is critical to achieve a parallel speedup and that the parameters of the model will give either conservative or optimistic an advantage.
\paragraph*{Allocation}\label{PholdTreeallocation}
The PholdTree benchmark can be configured to use 2 allocation schemes: breadth first and depth first. 
The breadth first allocation scheme will result in kernels that will form a dependency chain with multiple branches, much like in the Queue model. 
Such a linear dependency chain can result in a parallel speedup as we demonstrated with the Queue model, but this is not always true as we will demonstrate in this section. 
A single kernel that has an unbalanced number of atomic models or models with unequal computation load in transition functions will slow down the remainder of the chain. This effect is also apparent if the OS unfairly schedules the thread a kernel runs on. In conservative this will lead to excessive polling of the other kernels' eot values, whereas in optimistic this will lead to a cascade of reverts since dependent kernels will simulate ahead of the slower kernel. In Figures \ref{fig:pholdtree_visualize_parBFS} and \ref{fig:pholdtree_visualize_parDFS} the simulation trace is visualized for both allocation schemes highlighting the remarks made in this paragraph.

@@

Adevs' conservative implementation cannot handle the uncertainty (lack of non $\epsilon$ lookahead) here and is omitted from the comparison. Its parallel performance is two orders of magnitude slower than the sequential implementation. Upon investigation we conclude that this slowdown is caused by not implementing the \{begin/end\}lookahead() functions, which when not overridden in a model trigger an exception on each invocation. \\
This exception handling completely stalls the kernel. We do not implement this function in our PholdTree variant for adevs since this would force us to implement state saving in the model code and not use the provided state saving functionality in the kernel, as is done in dxex or PythonPDEVS's optimistic.\\
In dxex state saving is never required for a conservative simulation, this is handled transparently for the user who need not implement this behaviour. 
By using the state saving technique inside the model code we feel we would no longer compare identical models across different synchronization algorithms, although we do not doubt adevs' increased performance when these functions are overridden.
\\

@@

% Conservative
Conservative has a more nuanced speedup behaviour in this benchmark. We have detailed how a conservative kernel in dxex scales linear in the number of atomic models (if lookahead is $\epsilon$), this effect is more clearly visible here. \\
It is clear that as the fanout (n) of the model increases the kernel performance degrades rapidly. The intersection of the performance graph of each configuration is shifted to the right, the actual 1-speedup point increases as the kernelcount increases. For a d=4, n=5 configuration with 4 kernels each kernel has a load of ~1300 atomic models when it crosses the 1-speedup line. \\
An 8 kernel configuration with d=4 n=6 can sustain a load of ~2700 atomic models before it reaches that point. From the results we see that the 32-kernel configuration is not yet slowing down with the current parameters. \\
The modelcount is only a part of the explanation of this effect, as the kernelcount increases the amount of inter kernel messages will be split into more distinct sets which can be handled with more concurrency in dxex's architecture. 
That this effect is not always true can be observed from the d=4 n=3 32-kernel datapoint. Note that in this configuration 341 atomic models are distributed over 32 kernels, a relative low model load can more easily highlight synchronization overhead, even if allocation is optimal.\\
As with optimistic we do not include breadth first allocated benchmarks in this speedup plot since none of those achieve a speedup higher than 1.
\\
If we combine both to determine which synchronization protocol is more optimal we see in Figure \ref{fig:PholdTree_plot_weakall} that for these configurations conservative with 8 kernels is an ideal configuration up until n=5, where the modelcount starts to degrade performance. Optimistic with the same kernelcount is almost insensitive to the increase in modelcount and is therefore a more robust choice. Note how optimistic and conservative with 32 kernels at this point still have a reasonable speedup which is surprising given the synchronization overhead in such a large set of kernels.
\begin{figure}
	\center

@@

\subsection{Measuring Synchronization Overhead}
The simulation kernels have a non trivial overhead compared to a sequential kernel. In this section we will try to isolate this overhead. 
Profiling indicates that there are in the most trivial of models only 2 causes for computational load outside of the actual simulation : memory management and random number generation. By reducing the cost of both we will isolate the synchronization overhead. We use the Queue benchmark since it is highly parallel and its computational load in transitioning is only determined by random number generation and event/state allocation.
\paragraph*{Memory management}\label{sec:4-subsec:overhead-pgraph:memory}
\begin{figure}
	\center

@@

A sequential simulation kernel no allocated event will persist beyond a single time advance,  this enables the use of an arena-style allocator. Conservative and optimistic simulation need to use generic pool allocators since events are shared across kernels and thus have a different lifetime. \\

@@
